---
title: "tinytable in reveal.js: width fix"
subtitle: "Symptom → root cause → fix → recipe"
author: "記録: 伊藤成朗"
date: "2026-06-18"
format:
  revealjs:
    standalone: true
    embed-resources: true
    slide-number: true
    width: 1600
    height: 900
    theme: simple
    code-line-numbers: false
execute:
  echo: true
  eval: true
  warning: false
  message: false
---

```{r setup, echo=FALSE}
library(data.table); library(tinytable)
dt <- data.table(
  工程 = c("デザイン・仮説", "データ収集・整理", "推敲"),
  人間 = c("作り出す", "指示を出す", "監修、編集"),
  AI   = c("理論枠組み、仮説、データの整合性チェック",
           "指示通りに作業",
           "推計結果を表にして簡単な文章にする")
)
```

## Symptom

Table renders **full-width** — `tt(dt, width = 1)` (scalar `1`):

```{r symptom-table}
tb <- tt(dt, width = 1)   # scalar 1 → inline: table-layout:fixed; width:100% !important
tb <- style_tt(tb, j = 1:3, align = "l", fontsize = .8)
tb <- style_tt(tb, i = 0, align = "c")
tb <- theme_html(tb, class = "tinytable tinytable-hover")
tb
```

---

## Root cause

`tt()` injects table width directly into an **inline style** on `<table>`.
What each `width` arg produces:

```html
<!-- width = 1 (scalar): full-width, unkillable -->
<table style="table-layout: fixed; width: 100% !important; ...">

<!-- width = c(.2, .15, .4) (vector, sum=0.75): controllable -->
<table style="table-layout: fixed; width: 75%; ...">

<!-- width = NULL (no arg): shrinks to content -->
<table style="width: auto; ...">
```

. . .

The scalar `1` case writes **`!important` in the inline style** — the highest possible CSS priority.
No stylesheet rule can override it.

---

## What was tried and why each failed

| Attempt | Why it failed |
|---------|--------------|
| `.reveal table.tinytable { width: auto !important; }` | inline `!important` wins; CSS `!important` cannot beat inline `!important` |
| `.reveal table.tinytable { width: fit-content !important; }` | same — inline `!important` wins |
| `theme_revealjs()` | does not change the inline style; same `100% !important` |
| `as_raw_html()` + `gsub` | `as_raw_html()` is a **{gt}** function — not in tinytable |

The inline `width: 100% !important` set by tinytable is the highest-priority CSS value possible.
CSS overrides from any stylesheet cannot beat it.

---

## The fix

Supply a **column-width vector** — tinytable writes `width: XX%` (no `!important`):

```r
# Before: scalar width = 1 → inline width: 100% !important  (unkillable)
tb <- tt(dt, width = 1)

# After: vector summing to 0.70 → inline width: 70%  (70% of slide)
tb <- tt(dt, width = c(.175, .175, .35))
#                     ↑     ↑     ↑
#                     sum = 0.70
```

. . .

Column **proportions preserved** (example — original code used `c(.2, .15, .4)`):

| Column | Original | Fixed | Share of table |
|--------|----------|-------|----------------|
| 工程 | 0.200 | 0.187 | 26.7% |
| 人間 | 0.150 | 0.140 | 20.0% |
| AI   | 0.400 | 0.373 | 53.3% |
| **sum** | **0.750** | **0.700** | |

---

## Result: 70% width

```{r fix-table, echo=FALSE}
tb2 <- tt(dt, width = c(.187, .140, .373))
tb2 <- style_tt(tb2, j = 1:3, align = "l", fontsize = .8)
tb2 <- style_tt(tb2, i = 0, align = "c")
tb2 <- theme_html(tb2, class = "tinytable tinytable-hover")
tb2
```

No CSS file needed. `tinytable.css` should be **empty**.

---

## Why the fix works

```
tt(dt, width = c(.175, .175, .35))
    ↓  sum = 0.70
    ↓  tinytable writes: style="table-layout: fixed; width: 70%; ..."
    ↓  no !important → theme CSS can still override if needed
    ✓  table renders at 70% of slide width

tt(dt, width = 1)   ← DON'T USE
    ↓  tinytable writes: style="table-layout: fixed; width: 100% !important; ..."
    ↓  inline !important = highest CSS priority; nothing can override
    ✗  table locked at 100% width
```

**Key rule:** always supply a column-width vector; `sum = desired fraction of slide width`.
Leave `tinytable.css` empty — no CSS overrides needed.

---

## Recipe

```r
T <- 0.70   # target width as fraction of slide width

# column proportions (any values — will be rescaled)
col_w <- c(.25, .25, .50)
col_w <- col_w / sum(col_w) * T   # rescale to sum to T

tb <- tt(dt, width = col_w)
tb <- style_tt(tb, j = seq_along(col_w), align = "l", fontsize = .8)
tb <- style_tt(tb, i = 0, align = "c")
tb <- theme_html(tb, class = "tinytable tinytable-hover")
tb
# Leave tinytable.css empty.
```
